Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2023-01-07 unimodular-covers-CombTheory-reviewing.tex
dealt with referee X

Oda \cite{Oda1997} posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
Following \cite{HaaseHof, Tsuchiya}, we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
%
A lattice polytope $Q$ is called \emph{smooth} if it is simple and the primitive edge directions at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
\end{enumerate}
\end{conjecture}

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$. \Giulia{RefY says to add the definition of Minkowski sum here, rather than have the definition in section 3; I think it would break up the flow of the introduction. Perhaps a compromise is to add a comment like, "where + denotes the Minkowski sum defined in Section 3". Then again, the Cayley sum is defined both later in the intro and in section 3. What do you think?}
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Thm.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is called ``$\N$-Minkowski summand'' there).

\medskip 
Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes,
with special attention to dimension $3$ (in dimension 2 is it straightforward that ever lattice polygon has unimodular triangulations). 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds (affine images of $3$-cubes) and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

Then, the inequalities $b\ge 1+a \ge 2$ imply:
\[
u:=(1,1,0)\in \conv(p_1p_2q_3) \subset T_4, \quad
v:=(0,-1,0)\in \conv(p_1p_2q_4) \subset T_3.
\]

Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.

@@

We call \emph{band} any region of the form $f^{-1}([\alpha,\beta])$ for some functional $f\in (\R^3)^*\setminus \{0\}$ and closed interval $[\alpha,\beta]\subset \R$.
We claim that any band containing $T$ must contain at least three of the $q_i$s. 
This claim implies that the parallelepiped $P$, which is the intersection of three bands, contains at least one of the $q_i$s and hence it fully contains the corresponding $T_i$.

To prove the claim, suppose that $q_1\not\in B:= f^{-1}([\alpha,\beta])$ for a certain band $B \supset T$. 
Without loss of generality, say $f(q_1)<\alpha$. Then the equalities $q_1+q_2=p_3+p_4$ and $q_1+p_1=q_2+p_2$ respectively give:
\begin{gather}
\label{eq:first}
f(q_2) = f(p_3+p_4-q_1) = f(p_3)+f(p_4)-f(q_1) > 2\alpha-\alpha=\alpha,\\
\label{eq:second}
f(q_2) = f(q_1+p_1-p_2) = f(q_1)+(f(p_1)-f(p_2)) < \alpha + (\beta-\alpha) = \beta,
\end{gather}
so that $q_2 \in B$.

Inequality \eqref{eq:first} also implies
\begin{equation}
\label{eq:third}
f(q_1) <  f(p_i) < f(q_2), \quad\text{ for $i=3,4$}.
\end{equation}

The translation by the vector $\frac12 (p_1+p_2-p_3-p_4)$ sends $q_1,q_2,p_3,p_4$ to $p_2,p_1,q_4,q_3$ (in this order). By applying this to inequality \eqref{eq:third}, we obtain
\[
\alpha \le f(p_2) < f(q_i) < f(p_1) \le \beta, \quad\text{ for $i=3,4$},
\]
so that $q_3,q_4 \in B$.
%
This concludes the proof of the claim, and of the lemma.

@@

\begin{corollary}
\label{coro:coverpara}
Let $T$ be an empty lattice tetrahedron contained in a lattice parallelepiped $P$. Then, $T$ can be covered by unimodular tetrahedra contained in $P$.
\end{corollary}
\giulia{I would ignore comment Y.16, where if I understand correctly a different proof idea is suggested...}
\begin{proof}
We proceed by induction on the (normalized) volume of $T$, which is a positive integer. If this volume equals $1$ then $T$ is unimodular and there is nothing to prove, so we assume $T$ is not unimodular. Let $p_1, p_2, p_3, p_4$ denote the vertices of $T$.

@@

\begin{example}[A smooth 3-polytope without the circumscribed parallelepiped property]
 Let $P$ be the Cayley embedding of a long horizontal rectangle and a long vertical rectangle. That is,
$P=\conv([0,a]\times [0,1] \times\{0\} \cup [0,1]\times [0,b] \times\{1\})$, for big $a$ and $b$. This is smooth and contains the big empty tetrahedron $T$ with vertices (0,0,0), (a,1,0), (0,0,1), (1,b,1) which occupies most of its volume. In particular, none of the corner tetrahedra of $T$ is contained in $P$. Another way to see this is to write explicitly the remaining vertices $q_i$ of the circumscribed parallelepiped, which are $(\frac{a+1}{2}, \frac{b+1}{2}, 1), (\frac{a+1}{2}, \frac{b+1}{2}, 0), (\frac{1-a}{2}, \frac{b-1}{2}, 1), (\frac{a-1}{2}, \frac{1-b}{2}, 1)$. None of these points are contained in $P$, and therefore none of the corner tetrahedra are either.
\end{example}\giulia{Added the last part to answer comment Y.19}

\begin{example}[Non-IDP octahedron and triangular prism]
\label{ex:non-IDP}
The following lattice octahedron $Q$ and triangular prism $P$ are not IDP:
\begin{gather}
Q= \conv((0,1,1),(1,0,1),(1,1,0),(0,-1,-1),(-1,0,-1),(-1,-1,0)),\\
P=\conv((0,1,1),(1,0,1),(1,1,0),(-1,0,0),(0,-1,0),(0,0,-1)).
\end{gather}

Indeed, in both polytopes the only lattice points are the six vertices and the origin. The point $(1,1,1)$ lies in the second dilation but is not the sum of two lattice points in the polytope. Hence, they are not IDP, which implies they do not admit unimodular covers.
\end{example}

An affirmative answer to the following question (weaker than the circumscribed parallelepiped property) would still  imply that smooth $3$-polytopes can be unimodularly covered, and hence~\Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three is:

@@

The so-called \emph{Cayley Trick} is the isomorphism
\[
2\cayley(P, Q) \cap (\R^d\times \{1\}) \cong P+Q,
\]
which easily implies:

\begin{proposition}[see, e.g.~\protect{\cite[Thm.~0.4]{Tsuchiya}}]
\label{prop:mixedIDP}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is mixed IDP.
\end{proposition} \giulia{We just called the pair IDP in the introduction... Which notation do we prefer?}

The Cayley Trick also provides the following canonical bijections:

\[
\begin{array}{ccc}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow& \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow& \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow& \text{unimodular prod-simplices in $P + Q$}.
\end{array}
\]
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from any vertex of $T_1$ and from any vertex of $T_2$ form a unimodular basis.

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. A triangulation of $\cayley(P,Q)\subset \R^3$ consists of tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$, which are Cayley sums of a triangle in $P$ and a point in $Q$, or viceversa, are automatically unimodular. The case that we need to study are therefore tetrahedra of type $(2,2)$, which are Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$. These correspond to prod-simplices of two segments in $P+Q$, which are parallelograms.
The following lemma, whose proof we postpone to \Cref{sec:the_lemma}, is crucial to understand how to unimodularly cover these tetrahedra. 
We use the following conventions: if $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open line segments with endpoints $a,b$. Given a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and  the line spanned by $\vec s$ by $\vecline s$.

\begin{lemma}
\label{lemma:cayley}
%\paco{$Q$ is 2-dim, but $P$ may not}
Let $Q$ be a two-dimensional lattice polytope and $P$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. See \Cref{fig:strips}.
\end{lemma}

@@

Let us denote $\tilde u$, $\tilde p_1$, $\tilde p_2$, $\tilde q_1$, $\tilde q_2$ the points corresponding to $u, p_1, p_2, q_1, q_2$ in $\cayley(P,Q)$.
That is, $\tilde p_i = p_i\times\{1\}$, $\tilde q_i = 	q_i\times\{0\}$, and $\tilde u = u\times\{1\}$.\giulia{to match the figure, P is now at height 1 instead of 0}
Observe that the assumption $u\in((p_1, p_2) + \vecline q$ implies that one of the segments $[u,q_i]$ crosses one of the triangles $\conv(p_1,p_2,q_j)$, where $\{i,j\}=\{1,2\}$, see \Cref{fig:flip}.

@@

In turn, this means that the polytope $\conv(\tilde u, \tilde p_1, \tilde p_2, \tilde q_1, \tilde q_2) = \cayley(t,q)$
 has the following two triangulations:
\begin{gather*}
\mathcal T^+:= \left\{ \cayley(p,q), \cayley(t, \{q_j\}) \right\},
\\
\mathcal T^-:= \{ \cayley([p_1,u],q), \cayley([p_2,u],q), \cayley(t, \{q_i\}) \}.
\end{gather*}
The tetrahedra $\cayley(t, \{q_j\})$ and $\cayley(t, \{q_i\})$ are unimodular, which implies that $T=\cayley(p,q)$ has volume equal to the sum of the volumes of $\cayley([p_1,u],q)$ and $\cayley([p_2,u],q)$. In particular, we have covered $T$ by the three tetrahedra in $\mathcal T^-$, which are of smaller volume and hence have unimodular covers by induction hypothesis.
\end{proof}

@@

We can assume that $q_1$ is the origin, and thus $f_q(q)=0$. Observe that in the strip $q +\vecline p$, there is a unique lattice point on the line $f_q(x)=-1$;  indeed, since $q$ is primitive, the only way that in the strip there could be two lattice points on $f_q(x)=-1$ is if they were on the boundary of the strip, which would however imply that $p+q$ is a unimodular paralellogram, against our assumptions.
Since translating the polytopes independently by lattice vectors will not result in any loss of generality, we can assume that $p_1$ is that unique lattice point. That is, $f_q(p_1)=-1$, or equivalently, the triangle $\conv(q_1, q_2, p_1)$ is unimodular. Similarly, the unique lattice point in the strip on the line $f_q(x)=1$ is then $q_1+q_2 -p_1$.

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq f_p(p)\}$ and $V_2=\{f_p(x) \geq f_p(p)\}$. 
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants (see \Cref{fig:setup}). This is always possible via an affine transformation. \Giulia{I'm not sure this last sentence addition is necessary, but I thought it would help the ref clear up confusion (Y.31) about p being a vertical segment-sure, this isn't possible with a unimod transformation, but we don't need to fix the lattice...}
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$
Then:
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ). 
\]

@@

The same holds for $P$ and the strip $p+\vecline q$, and we call the segments $\ell=[l_1, l_2]$ and $r=[r_1, r_2]$, with $\ell$ and $r$ crossing the strip $p + \vecline q$  in $V_1$ and $ V_2$ respectively. The only difference is that in the case that $P$ is one dimensional we have $\ell=r=p$.  Again we have
\begin{gather}
\label{eq:widthp}
\begin{array}{cc}
f_q(l_1) \leq f_q(p_1), &
f_q(l_2) \geq f_q(p_2), \\
f_q(r_1) \leq f_q(p_1),&
f_q(r_2) \geq f_q(p_2).
\end{array}
\end{gather}

Observe that a priori one of $l$ and $r$ can coincide with $p$, if this is on the boundary of $P$, and similarly one of $t,b$ might be $q$, if this is on the boundary of $Q$.

@@

\begin{proof}
Both $t$ and $b$ must intersect $p$, as seen in the proof of \Cref{claim:b_and_t}.  Their intersections with $p$ are thus endpoints of a segment of width with respect to $f_q$ less than $w$, the width of $p$. Since $t$ and $b$ cross in $V_2$, the same is true for any segment parallel to $p$ contained in $Q \cap V_2$.

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption on $p_1$, the values of $g$ on these segments differer by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]
\giulia{changed the +1 to a -1, so that the inequalities in the following claim are the right way around--we had flipped everything}

>>>>

Oda \cite{Oda1997} posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
Following \cite{HaaseHof, Tsuchiya}, we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d,
\end{align*}
where $A+B :=\{a+b: a\in A, b\in B\}$ denotes the \emph{Minkowski sum} of two sets $A,B\subset \R^d$.

A lattice polytope $Q$ is called \emph{smooth} if it is simple and the primitive edge directions at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
\end{enumerate}
\end{conjecture}

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$. 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Thm.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is called ``$\N$-Minkowski summand'' there).

\medskip 
Motivated by these and other questions, several authors have studied the IDP property for different classes of lattice polytopes,
with special attention to dimension $3$ (in dimension 2 it is straightforward that every lattice polygon has unimodular triangulations). 
%
For example,  very recently
Beck et al.~\cite{BHHHJKM2019} proved that all smooth centrally symmetric $3$-polytopes are IDP.
More precisely, they show that any such polytope can be covered by lattice 
parallelepipeds (affine images of $3$-cubes) and unimodular simplices, both of which are trivially IDP.
%
In \Cref{sec:parallelepipeds} we show:

@@

Then, the inequalities $b\ge 1+a \ge 2$ imply:
\[
u:=(1,1,0)\in \conv(p_1,p_2,q_3) \subset T_3, \quad
v:=(0,-1,0)\in \conv(p_1,p_2,q_4) \subset T_4.
\]

Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1,q_4,p_2,q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2,p_3,q_1,p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2,p_3,q_1,p_4)$ must contain lattice points in both triangles $\conv(q_2,p_3,p_4)\subset T_1$ and $\conv(q_1,p_3,p_4)\subset T_2$.

@@

We call \emph{band} any region of the form $f^{-1}([\alpha,\beta])$ for some linear functional $f\in (\R^3)^*\setminus \{0\}$ and closed interval $[\alpha,\beta]\subset \R$.
We claim that any band containing $T$ must contain at least three of the $q_i$s. 
This claim implies that the parallelepiped $P$, which is the intersection of three bands, contains at least one of the $q_i$s and hence it fully contains the corresponding $T_i$.

To prove the claim, suppose that $q_1\not\in B:= f^{-1}([\alpha,\beta])$ for a certain band $B \supset T$. 
Without loss of generality, say $f(q_1)<\alpha$. Then the equalities $q_1+q_i=p_j+p_k$ and $q_1+p_1=q_i+p_i$, where $\{i,j,k\}=\{2,3,4\}$, respectively give:
\paco{used $i$ instead of $2$, to shorten proof}
\begin{gather}
\label{eq:first}
f(q_i) = f(p_j+p_k-q_1) = f(p_j)+f(p_k)-f(q_1) > 2\alpha-\alpha=\alpha,\\
\label{eq:second}
f(q_i) = f(q_1+p_1-p_2) = f(q_1)+f(p_1)-f(p_i) < \alpha + \beta-\alpha = \beta,
\end{gather}
so that $q_i \in B$ for $i\in \{2,3,4\}$.
%
%Inequality \eqref{eq:first} also implies
%\begin{equation}
%\label{eq:third}
%f(q_1) <  f(p_i) < f(q_2), \quad\text{ for $i=3,4$}.
%\end{equation}
%
%The translation by the vector $\frac12 (p_1+p_2-p_3-p_4)$ sends $q_1,q_2,p_3,p_4$ to $p_2,p_1,q_4,q_3$ (in this order). By applying this to inequality \eqref{eq:third}, we obtain
%\[
%\alpha \le f(p_2) < f(q_i) < f(p_1) \le \beta, \quad\text{ for $i=3,4$},
%\]
%so that $q_3,q_4 \in B$.
%
This concludes the proof of the claim, and of the lemma.

@@

\begin{corollary}
\label{coro:coverpara}
Let $T$ be an empty lattice tetrahedron contained in a lattice parallelepiped $P$. Then, $T$ can be covered by unimodular tetrahedra contained in $P$.
\end{corollary}
%\giulia{I would ignore comment Y.16, where if I understand correctly a different proof idea is suggested...}
\begin{proof}
We proceed by induction on the (normalized) volume of $T$, which is a positive integer. If this volume equals $1$ then $T$ is unimodular and there is nothing to prove, so we assume $T$ is not unimodular. Let $p_1, p_2, p_3, p_4$ denote the vertices of $T$.

@@

\begin{example}[A smooth 3-polytope without the circumscribed parallelepiped property]
 Let $P$ be the Cayley embedding of a long horizontal rectangle and a long vertical rectangle. That is,
$P=\conv([0,a]\times [0,1] \times\{0\} \cup [0,1]\times [0,b] \times\{1\})$, for big $a$ and $b$. This is smooth and contains a big empty tetrahedron $T$ with vertices $(0,0,0)$, $(a,1,0)$, $(0,0,1)$, $(1,b,1)$ which occupies most of its volume. In particular, none of the corner tetrahedra of $T$ is contained in $P$.

More explicitly, the the remaining vertices $q_i$ of the circumscribed parallelepiped are $(\frac{a+1}{2}, \frac{b+1}{2}, 1), (\frac{a+1}{2}, \frac{b+1}{2}, 0), (\frac{1-a}{2}, \frac{b-1}{2}, 1), (\frac{a-1}{2}, \frac{1-b}{2}, 1)$. None of these points are contained in $P$, and therefore none of the corner tetrahedra are either.
\end{example}
%\giulia{Added the last part to answer comment Y.19}

\begin{example}[Non-IDP polytopes with four facet directions]
\label{ex:non-IDP}
The following triangular prism $P$ and centrally symmetric octahedron $Q$ are not IDP:
\begin{gather}
P=\conv((0,1,1),(1,0,1),(1,1,0),(-1,0,0),(0,-1,0),(0,0,-1)),\\
Q= \conv((0,1,1),(1,0,1),(1,1,0),(0,-1,-1),(-1,0,-1),(-1,-1,0)).
\end{gather}

Indeed, in both cases the 
%only lattice points are the six vertices and the origin. The 
point $(1,1,1)$ lies in the second dilation but is not the sum of two lattice points in the polytope. 
%Hence, they are not IDP, which implies they do not admit unimodular covers.
\end{example}

An affirmative answer to the following question (weaker than the circumscribed parallelepiped property) would still  imply that smooth $3$-polytopes can be unimodularly covered and, hence,~\Cref{conj:Oda}\eqref{itm:smoothIDP} in dimension three:

@@

The so-called \emph{Cayley Trick} is the isomorphism
\[
2\cayley(P, Q) \cap (\R^d\times \{1\}) = (P+Q) \times \{1\} \cong P+Q,
\]
which easily implies:

\begin{proposition}[see, e.g.~\protect{\cite[Thm.~0.4]{Tsuchiya}}]
\label{prop:mixedIDP}
If $\cayley(P,Q)$ is IDP then the pair $(P,Q)$ is IDP.
\end{proposition} 
%\giulia{We just called the pair IDP in the introduction... Which notation do we prefer?}

The Cayley Trick also provides the following canonical bijections:
\[
\begin{array}{ccc}
\text{polyhedral subdivisions of $\cayley(P,Q)$} &\leftrightarrow& \text{mixed subdivisions of $P + Q$}\\
\text{triangulations of $\cayley(P,Q)$} &\leftrightarrow& \text{fine mixed subdivisions of $P + Q$}\\
\text{unimodular simplices in $\cayley(P,Q)$} &\leftrightarrow& \text{unimodular prod-simplices in $P + Q$}.
\end{array}
\]
See \cite{DLRS2010} for more details on the Cayley Trick and on triangulations and polyhedral subdivisions of polytopes.
In fact these bijections can be taken as definitions of the objects in the right-hand sides. In particular, 
we call \emph{prod-simplices} in $P+Q$ the Minkowski sums $T_1+T_2$ where $T_1\subset P$ and $T_2\subset Q$ are simplices with complementary affine spans. A prod-simplex is \emph{unimodular} if the edge vectors from any vertex of $T_1$ and from any vertex of $T_2$ form a unimodular basis.

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. A triangulation of $\cayley(P,Q)\subset \R^3$ consists of tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$, which are Cayley sums of an empty (hence unimodular) triangle in $P$ and a point in $Q$, or viceversa, are automatically unimodular. The case that we need to study are therefore tetrahedra of type $(2,2)$, which are Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$. These correspond to prod-simplices of two segments in $P+Q$, which are \emph{parallelograms}.
The following lemma, whose proof we postpone to \Cref{sec:the_lemma}, is crucial to understand how to unimodularly cover these tetrahedra.

We use the following conventions: if $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open line segments with endpoints $a,b$. Given a segment $s=[a,b]$, we denote $\vec s$ the vector $b-a$ and denote $\vecline s$ the line spanned by $\vec s$. A lattice parallelogram $p+q$ is called \emph{unimodular} if it is a fundamental domain for the lattice. Equivalently, if $\cayley(p,q)$ is a unimodular tetrahedron.
\paco{added def of unimodular paralellogram}

\begin{lemma}
\label{lemma:cayley}
\paco{referee X suggests to give an ``example'' of the lemma, but I am not sure what could be helpful}
Let $Q\subset \R^2$ be a two-dimensional lattice polytope and $P\subset \R^2$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. 
\end{lemma}

See \Cref{fig:strips} for an illustration of the two regions in the statement, which we call \emph{strips}. In this figure and the forthcoming ones in \Cref{sec:the_lemma} 
we draw $p$ as a vertical segment and $q$ as a horizontal one for convenience. This is always possible via a linear transformation (which of course changes the lattice; in the proof we do not assume the lattice to be $\Z^2$). 
\paco{moved this remark here, since it already affects this proof}
%\Giulia{I'm not sure this last sentence addition is necessary, but I thought it would help the ref clear up confusion (Y.31) about p being a vertical segment-sure, this isn't possible with a unimod transformation, but we don't need to fix the lattice...}

@@

Let us denote $\tilde u$, $\tilde p_1$, $\tilde p_2$, $\tilde q_1$, $\tilde q_2$ the points corresponding to $u, p_1, p_2, q_1, q_2$ in $\cayley(P,Q)$.
That is, $\tilde p_i = p_i\times\{1\}$, $\tilde q_i = 	q_i\times\{0\}$, and $\tilde u = u\times\{1\}$.\giulia{to match the figure, P is now at height 1 instead of 0}
Observe that the assumption $u\in((p_1, p_2) + \vecline q$ implies that one of the segments $[u,q_i]$ crosses the interior of one of the triangles $\conv(p_1,p_2,q_j)$, where $\{i,j\}=\{1,2\}$. Without loss of generality assume that $[\tilde{u},\tilde{q}_2]$ crosses $\conv(\tilde{p}_1,\tilde{p}_2,\tilde{q}_1)$, as in \Cref{fig:flip}.

@@

In turn, this means that the polytope $\conv(\tilde u, \tilde p_1, \tilde p_2, \tilde q_1, \tilde q_2) = \cayley(t,q)$
 has the following two triangulations:
\begin{gather*}
\mathcal T^+:= \left\{ \cayley(p,q), \cayley(t, \{q_2\}) \right\},
\\
\mathcal T^-:= \{ \cayley([p_1,u],q), \cayley([p_2,u],q), \cayley(t, \{q_1\}) \}.
\end{gather*}
The tetrahedra $\cayley(t, \{q_1\})$ and $\cayley(t, \{q_2\})$ are unimodular, which implies that $T=\cayley(p,q)$ has volume equal to the sum of the volumes of $\cayley([p_1,u],q)$ and $\cayley([p_2,u],q)$. In particular, we have covered $T$ by the three tetrahedra in $\mathcal T^-$, which are of smaller volume and hence have unimodular covers by induction hypothesis.
\end{proof}

@@

\paco{rphrased to address X.10}
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that $q_1$ is the origin and that $f_q(p_1)=-1$. 
Observe that then $f_q(p_2)$ must be strictly positive, since $f_q(p_2)=0$ would imply that $p+q$ is a unimodular paralellogram. Moreover, since $p$ is primitive $p_1$ and $p_2$ cannot be in the boundary of $q +\vecline p$, which implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

%
%;  indeed, since $q$ is primitive, the only way that in the strip there could be two lattice points on $f_q(x)=-1$ is if they were on the boundary of the strip, which would however imply that $p+q$ is a unimodular paralellogram, against our assumptions.
%Since translating the polytopes independently by lattice vectors will not result in any loss of generality, we can assume that $p_1$ is that unique lattice point. That is, $f_q(p_1)=-1$, or equivalently, the triangle $\conv(q_1, q_2, p_1)$ is unimodular. Similarly, the unique lattice point in the strip on the line $f_q(x)=1$ is then $q_1+q_2 -p_1$.

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq f_p(p)\}$ and $V_2=\{f_p(x) \geq f_p(p)\}$.%
\paco{added footnote}
\footnote{Since in our figures $p$ and $q$ are vertical and horizontal, we use the letters $V$ and $H$ for the half-planes they defined. Similarly, later in the proof we use the letters $b$, $t$, $r$, and $l$ for certain  points and segments meaning ``bottom'', ``top'', ``right'' and ``left''.}
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$.
Then:
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ). 
\]

@@

The same holds for $P$ and the strip $p+\vecline q$, and we call the segments 
\paco{not sure why we use $\ell$ for the segment and $l$ for the endpoints...}
$\ell=[l_1, l_2]$ and $r=[r_1, r_2]$, with $\ell$ and $r$ crossing the strip $p + \vecline q$  in $V_1$ and $ V_2$ respectively. The only difference is that in the case that $P$ is one dimensional we have $\ell=r=p$.  Again we have
\begin{gather}
\label{eq:widthp}
\begin{array}{cc}
f_q(l_1) \leq f_q(p_1), &
f_q(l_2) \geq f_q(p_2), \\
f_q(r_1) \leq f_q(p_1),&
f_q(r_2) \geq f_q(p_2).
\end{array}
\end{gather}

Observe that a priori one of $\ell$ and $r$ can coincide with $p$, if $p$ is on the boundary of $P$, and similarly one of $t,b$ might be $q$, if $q$ is on the boundary of $Q$.

@@

\begin{proof}
Both $t$ and $b$ must intersect $p$, as said in the proof of \Cref{claim:b_and_t}.  Their intersections with $p$ are thus endpoints of a segment of width with respect to $f_q$ less than $w$, the width of $p$. Since $t$ and $b$ cross in $V_2$, the same is true for any segment parallel to $p$ contained in $Q \cap V_2$.

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption on $p_1$, the values of $g$ on these segments differ by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]
\giulia{changed the +1 to a -1, so that the inequalities in the following claim are the right way around--we had flipped everything}
